---
title: "Reproducing Data Peer Assignment 1"
author: "J.Chris Holt"
date: "6/18/2017"
output: 
    html_document: 
        keep_md: true
---

## Load Libraries
```{r setup,echo=TRUE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(lubridate)
library(ggplot2)
```

## Loading and preprocessing the data
```{r read_file, echo=TRUE}
## Read in file and set POSIXct date
activity_raw <- read.csv(file = "activity.csv", sep = ",", stringsAsFactors = FALSE)
activity_raw$date <- ymd(activity_raw$date)
```

## What is mean total number of steps taken per day?
```{r mean_steps, echo=TRUE, warning=FALSE, fig.width=10}
# histogram of total steps taken
ggplot(data= activity_raw, aes(steps)) + geom_histogram() +
    ggtitle("Frequency of Steps Taken")
```

### The mean and median of total number of steps taken per day
```{r am_steps, echo=TRUE, warning=FALSE, fig.width=10}
apd <- mean(activity_raw$steps, na.rm = TRUE)
mpd <- median(activity_raw$steps, na.rm = TRUE)
```
  
* Mean: `r apd`  
* Median: `r mpd`

## What is the average daily activity pattern?

```{r daily_pattern, echo=TRUE, warning=FALSE, fig.width=10}
avg_perint <- activity_raw %>%
    group_by(interval) %>%
    summarize(avg = mean(steps, na.rm=TRUE))

ggplot(data = avg_perint, aes(interval, avg)) +
    geom_line() + ggtitle("Average Daily Steps Taken") +
    xlab("5 Minute Interval") + ylab("Average Steps")
```

### The busiest 5-minute-interval
```{r daily_pattern_avg, echo=TRUE, warning=FALSE, fig.width=10}
max5 <- avg_perint %>% filter(avg == max(avg_perint$avg)) %>% select(interval)
```
The most active 5-minute-interval across all the days in the dataset is `r max5`.

### Imputing Missing Values
Missing steps values are imputed by using the average 5-minute-interval calculated in the prior step.
```{r imputing, echo=TRUE, warning=FALSE, fig.width=10, message=FALSE}

act <- activity_raw
act$steps <- ifelse(is.na(act$steps) == TRUE, 
                    avg_perint$avg[avg_perint$interval %in% act$interval], 
                    act$steps)

# histogram of total steps taken
ggplot(data= act, aes(steps)) + geom_histogram() +
    ggtitle("Frequency of Steps Taken with Imputed Values")
```

### The mean and median of total number of steps per day with imputed values
```{r am_steps_imputed, echo=TRUE, warning=FALSE, fig.width=10}
apdi <- mean(act$steps)
mpdi <- median(act$steps, na.rm = TRUE)
```
The mean and median have not changed after the missing steps values are imputed.
  
* Mean: `r apdi`  
* Median: `r mpdi`

## Are there differences in activity patterns between weekdays and weekends?

```{r activity_weekdays, echo=TRUE, warning=FALSE, fig.width=10}
act_weekday <- act %>%
    mutate(day_type = as.factor(ifelse(weekdays(act$date) %in% c("Saturday","Sunday"), "weekend","weekday"))) %>%
    group_by(interval, day_type) %>%
    summarize(avg = mean(steps, na.rm=TRUE))

ggplot(data = act_weekday, aes(interval, avg)) +
    geom_line() + facet_grid(day_type~.) +
    ggtitle("Average Daily Steps Taken") +
    ylab("Steps Taken") +
    xlab("5-Minute-Interval")
```
